# Setup {.unnumbered}  

## Installing and Using Required Packages {.unnumbered}  
Throughout this tutorial, we’ll use a few essential R packages to manipulate data, run models, and create plots.   
Below are the core packages, what they do, and how to install them.  

Packages we’ll use:  

**ggplot2** – For plotting (e.g., scatterplots, regression lines)    
**dplyr** – For data wrangling (filtering, mutating, summarizing, etc.)     
**gridExtra** – To combine multiple plots into one figure    
**stats** – Comes with base R and used for regression (lm())    
**pacman** – Simplifies package management in R
**broom (optional)** – Makes model summaries easier to work with    

```{r}

# install.packages(c("ggplot2", "dplyr", "gridExtra", "pacman", "broom"))

```

## Loading Packages {.unnumbered}  

```{r}

library(ggplot2)
library(dplyr)
library(gridExtra)

# Optional if using tidy model outputs:
# library(broom)

```

## Difference Between install.packages() and library() {.unnumbered}

- install.packages("dplyr") downloads and installs the package — you only need to do this once per computer.  
- library(dplyr) loads the package into your R session — you need to run this each time you use it.  

## Session Information {.unnumbered}

It's always good practice to include your session info at the end of your analysis. This gives a snapshot of:  

- Your R version and system details  
- All the packages that were loaded  
- The versions of those packages  

This is especially useful when:  
- You're debugging errors  
- You're submitting assignments  
- You're collaborating with others  

## Why include this? {.unnumbered}

Sometimes code behaves differently depending on the version of a package or even the version of R itself. Including your session info makes your work **reproducible and easier to troubleshoot**.

## Base R version (simple) {.unnumbered}

This function comes with R and gives you basic session details.

```{r}

sessionInfo()

```
